---
title: "Big Countries"
output: html_notebook
---

## Country categorization

```{r}
library(tidyverse)
source('mix.R')
source('population.R')
source('plot_utils.R')
```

### EU countries
```{r}
eu_countries <- read.csv('data/eu_countries.csv')

eu <- countries |>
  inner_join(eu_countries, join_by(iso_3digit_alpha == iso_3digit_alpha)) |>
  mutate(big_category = 'EU')
eu
```

### Big countries

Twelve biggest countries not in the EU

```{r}
big <- countries |>
  anti_join(eu_countries, join_by(iso_3digit_alpha == iso_3digit_alpha)) |>
  slice_max(population, n=12) |>
  mutate(big_category=iso_3digit_alpha)
big
```

### World

World population

```{r}
world_pop <- sum(countries$population)
world_pop
```

Categorize countries as either a big country themselves, part of the EU, or part of the world.

```{r}
categorized <- countries |>
  left_join(union(big, eu) |> select(iso_3digit_alpha, big_category), join_by(iso_3digit_alpha == iso_3digit_alpha)) |>
  arrange(gdp_per_capita) |>
  mutate(
    gdp = gdp_per_capita * population,
    big_category = coalesce(big_category, 'World'),
    p = (cumsum(as.numeric(population)) - population/2)/world_pop
  )
```

Sort categories by their average GDP/capita

```{r}
category_order <- categorized |>
  mutate(
    big_name = ifelse(big_category == iso_3digit_alpha, name, big_category)
  ) |>
  group_by(big_category, big_name) |>
  summarise(
    gdp=sum(gdp),
    population=sum(population)
  ) |>
  mutate(
    gdp_per_capita = gdp/population,
    ) |>
  arrange(-gdp_per_capita)
category_order
```

Put categories in that order for plots by creating a factor in that order

```{r}
categorized$big_category <- factor(categorized$big_category, levels=category_order$big_category )
```

## Mixture of countries

Mix countries together with

* a mean of their log gdp/capita
* a standard deviation of a log-normal distribution with gini coefficient equal to the country's income inequality gini coefficient
* weight proportional to their population

```{r}
m <- log(categorized$gdp_per_capita)
s <- 2*erf.inv(categorized$gini)
w <- categorized$population/sum(categorized$population)

world <- mix(mapply(Norm, m, s), w)
```

This distributes the population of each country among world population based on GDP inequality between countries and income inequality within countries.

```{r, fig.width=12}
resolution <- 1024

ds <- world@quantile_bins(resolution)
df <- reshape2::melt(t(ds)) |>
  mutate(
    p = (Var1-0.5)/resolution,
    big_category = categorized[Var2,'big_category'])

ggplot(df, aes(x=p, y=value, fill=big_category)) + 
  geom_bar(stat="identity", width=1/resolution) +
  scale_fill_manual(values = rev(c("#b98d32", "#115740", "#1B7339", "#FF9933", "#815f4b", "#ff8080", "#76a442", "#CCCCCC", "#ab8179", "#dcacb1", "#EE1C25", "#001489", "#BC002D", "#d9d2d6")),
    labels = category_order$big_name
  ) +
  labs(
    title = "World population by national GDP/capita, 2021",
    subtitle = "mixing countries by log-normal GDP/capita, income inequality, and population",
    y = "portion of population", x = "mixed GDP/capita quantile",
    fill="Country")
```

The area of the graph is the population of the world. Mostly China 🇨🇳 in red and India 🇮🇳 in orange.

The population is sorted by national GDP/capita increasing from left to right across the graph, with the GDP for each country blurred by that nation's income inequality.

In the other direction the 12 biggest countries and the EU are sorted bottom to top by increasing GDP/capita, with groups for the EU and all of the rest of the world.

## GDP/capita regression

Estimate GDP/capita based on the mixture of countries as the weighted geometric mean of each country GDP/capita.

```{r}
log_gdp_regression <- t(m) %*% ds
df <- reshape2::melt(t(log_gdp_regression)) |>
  mutate(
    p = (Var1-0.5)/resolution,
    gdp_per_capita = exp(value)) |>
  select(p, log_gdp_per_capita=value, gdp_per_capita)

gdp_by_quantile <- data.frame(
  categorized$iso_3digit_alpha,
  p=world@p(m),
  log_y=m,
  log_y_hat=t(m %*% scale1(world@d(m))),
  gdp_per_capita=categorized$gdp_per_capita
)
gdp_by_quantile
```


```{r}
r2 <- r_squared(gdp_by_quantile$log_y_hat, gdp_by_quantile$log_y)

ggplot() + 
  geom_line(data=df, aes(x=p, y=gdp_per_capita, color="Estimated")) +
  geom_point(data=gdp_by_quantile, aes(x=p, y=gdp_per_capita, color="Countries"), shape=3, size=0.75) + 
  scale_y_log10(labels = scales::dollar) +
  geom_text( aes(x=0, y=max(gdp_by_quantile$gdp_per_capita), hjust=0, vjust=1), label = r2_label(r2), parse = TRUE, show.legend = FALSE) +
  scale_color_manual(
    name = NULL,
    values = c("black", "#3399DD"),
    guide = guide_legend(override.aes = list(
      linetype = c(0, 1),
      shape = c(3, NA)
    ))
  ) + 
  labs(title = "Estimated GDP per capita", subtitle="mixing countries by log-normal GDP/capita, income inequality, and population", y="GDP per capita", x="mixed GDP/capita quantile")

```


